---
title: "US Mass Shooting"
author: "Qixing Li"
date: "January 2, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, warning=FALSE)
```

##1.Introduction
As the most mass shooting country in the world, United States suffered at least 15 mass shooting in the past decade. According to Wikipedia, 305 people died and more than 1100 were injured in those shooting. On the night of October 1, 2017, a gunman opened fire on a crowd attending the final night of a country music festival in Las Vegas, killing 58 people and injuring more than 800. This shooting is the worst mass shooting in modern history of the United States.

The gun control issue has always been debated for a long time, probably ever since they were invented.However, the project isn't about arguing the gun control issue or Second Amendment, the purpose of the project is only about to gain some statistic insights on these mass shooting incidents and get a better understanding of the relationships between characteristics of the shooters and the incidents by performing some spatial and statistical analysis on the mass shooting dataset from [Kaggle](https://www.kaggle.com/zusmani/us-mass-shootings-last-50-years). The dataset is also avaiable in my [Github](https://github.com/qixing810/Kaggle-US-Mass-Shootings).


##2.Methodology
**1.Load all the required libraries**
```{r}
library(data.table)
library(leaflet)
library(lubridate)
library(magrittr)
library(maps)
library(plotly)
library(stringr)
library(tidyverse)
```

**2.Load & take a glimpse of the dataset**
```{r}
shoot <- read.csv("C:/SPS/projects/Mass Shootings Dataset Ver 5.csv",header=TRUE,stringsAsFactors=FALSE)
head(shoot)
glimpse(shoot)
summary(shoot)
```

**3.Reconstruct the dataset**
**3.1 Sepearte the *Location* into *City* and *State* two variables**
```{r}
shoot <- shoot %>% separate(Location,into = c("City","State"), sep = ", ",remove = FALSE)
```

From the warning messages, almost 15% of the samples in the column *Location* contain NA and the location of the incidents is not difficiult to figure out. So manually fill out those NA seems necessary. 
```{r}
shoot1 <- read.csv("C:/SPS/projects/Mass Shootings Dataset fixed.csv",header=TRUE,stringsAsFactors=FALSE)
shoot1 <- shoot1 %>% 
  separate(Location,into = c("City","State"), sep = ", ", remove = FALSE)
```

Abbreviations needed to be repleace by names of states.
```{r}
pattern <- c("CA|CO|LA|MD|NV|PA|TX|WA|VA")
replacement <- c("California","Colorado","Louisiana","Maryland","Nevada","Pennsylvania","Texas","Washington","Virginia")
shoot1$State <- shoot1$State %>%
  str_replace_all(pattern = pattern, replacement = replacement)
shoot1$State <- shoot1$State %>%
  str_replace_all(c("Texas "="Texas"," Virginia"="Virginia"))
```

**3.2 Extract year and month from column *Date* **
```{r}
shoot1 <- shoot1 %>% 
  mutate(Date=mdy(shoot1$Date),year=year(Date))
shoot1 <- shoot1 %>% 
  mutate(month=month(Date))
shoot1 <- shoot1 %>%
  mutate(decade = case_when(year >=1960 & year<1970 ~ "1960s",
                            year >=1970 & year<1980 ~ "1970s",
                            year >=1980 & year<1990 ~ "1990s",
                            year >=1990 & year<2000 ~ "1990s",
                            year >=2000 & year<2010 ~ "2000s",
                            year >=2010 & year<2020 ~ "2010s"))
```

**3.3 Deal with column *Mental.Health.Issues* **
```{r}
shoot1$Mental.Health.Issues <- if_else(shoot1$Mental.Health.Issues=="unknown","Unknown",shoot1$Mental.Health.Issues)
```

**3.4 Deal with column *Race* **
```{r}
shoot1$Race <- if_else(str_detect(shoot1$Race,"Black American or African American"),"Black",shoot1$Race)
shoot1$Race <- if_else(str_detect(shoot1$Race,"White American or European American"),"White",shoot1$Race)
shoot1$Race <- if_else(str_detect(shoot1$Race,"Asian American"),"Asian",shoot1$Race)
shoot1$Race <- if_else(shoot1$Race == "Some other race","Other",shoot1$Race)
shoot1$Race <- if_else(shoot1$Race == "Two or more races","Other",shoot1$Race)
shoot1$Race <- if_else(shoot1$Race == "Native American or Alaska Native","Native",shoot1$Race)
shoot1$Race <- if_else(shoot1$Race == "","Other",shoot1$Race)
shoot1$Race <- str_to_upper(shoot1$Race)
```

**3.5 Column *Gender* abbreviations needed to be repleace **
```{r}
shoot1$Gender <- if_else(shoot1$Gender=="M","Male",shoot1$Gender)
shoot1$Gender <- if_else(shoot1$Gender=="F","Female",shoot1$Gender)
shoot1$Gender <- if_else(shoot1$Gender=="M/F","Male/Female",shoot1$Gender)
```

**3.6 Deal with column *Cause* **
```{r}
shoot1$Cause <- if_else(shoot1$Cause=="","unknown",shoot1$Cause)
```

**3.7 Deal with column *Age* **
```{r}
temp <- shoot1 %>%
  separate_rows(Age,sep=",") %>%
  select(Age) %<>%
  mutate(Age = cut(as.integer(Age),breaks = c(10,20,30,40,50,60,70),
                   labels=c("10-20","20-30","30-40","40-50","50-60","60-70"))) %>%
  na.omit()
```

**4 Exploratory Data Analysis**
**4.1 Analyze the trend by year**
**Victims by year**
```{r}
victims_year <- shoot1 %>%
  group_by(year) %>%
  summarise(total=sum(Total.victims)) %>%
  ggplot(aes(x=year,y=total))+
  geom_bar(stat = 'identity',fill='blue')+
  labs(title = "US Mass Shooting Victims from 1966 to 2017",
       xlab = "year", ylab = "Number of Victims")
ggplotly(victims_year)
```

**Incidents by year**
```{r}
incidents_year <- shoot1 %>%
  group_by(year) %>%
  count() %>%
  ggplot(aes(x=year,y=n))+
  geom_bar(stat = 'identity',fill='blue')+
  labs(title = "US Mass Shooting Incidents from 1966 to 2017")+
  xlab("Year")+
  ylab("Number of Incidents")
ggplotly(incidents_year)
```

**Incidents by month**
```{r}
incidents_month <- shoot1 %>%
  group_by(month) %>%
  count() %>%
  ggplot(aes(x=factor(month),y=n))+
  geom_bar(stat = 'identity',fill='blue')+
  labs(title = "Incidents happened by month ")+
  xlab("Month")+
  ylab("Number of Incidents")
ggplotly(incidents_month)
```

**4.2 Analyze the distribution of the incidents **
**The 10 states have the most incidents**
```{r}
incidents_states <- shoot1 %>%
  group_by(State) %>%
  count() %>%
  arrange(desc(n)) %>%
  head(10) %>%
  ggplot(aes(x=reorder(State,n),y=n))+
  geom_bar(stat = 'identity',aes(fill = State))+
  labs(title = "The 10 states have the most incidents")+
  xlab("State")+
  ylab("Number of Incidents")+
  theme(legend.position='none')+
  coord_flip()
ggplotly(incidents_states)
```

**Analyze incidents in maps**
```{r}
incid_states <- shoot1 %>%
  group_by(State) %>%
  count() 
colnames(incid_states)<-c("region","Incidents")
incid_states$region <- str_to_lower(incid_states$region)

states <- map_data("state")  

comb <- states %>%
  left_join(incid_states,by="region")

incident_map <- comb %>% 
  ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group,fill = Incidents))+ 
  geom_point(data=shoot1,aes(x=Longitude,y=Latitude,size=Total.victims),color='red',alpha=0.5)+
  xlim(-130,-65)+
  ylim(25,50)
ggplotly(incident_map)
```

**Further analyze total victims in maps**
```{r}
shoot1 %>%
  leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  fitBounds(-124,30,-66,43) %>%
  addCircles(color = "blue",lng = ~Longitude,lat = ~Latitude,weight = 1,
             radius = ~sqrt(Total.victims) * 25000,popup = ~Summary)
  
```

**Click to see details in maps**
```{r}
shoot1 %>%
  leaflet() %>%
  fitBounds(-124,30,-66,43) %>%
  addProviderTiles(providers$CartoDB.DarkMatter, group="Dark") %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addLayersControl(baseGroups=c('Dark','Light')) %>%
  addTiles() %>%
  addMarkers(~Longitude, ~Latitude, 
             clusterOptions = markerClusterOptions(),
             popup = ~Summary,
             label = ~Location)
```


**Analyze the characters of the shooters**
**Race of the shooter**
```{r}
shoot1 %>%
  group_by(Race) %>%
  summarise(total=sum(Total.victims)) %>%
  plot_ly(labels= ~Race,values= ~total, type="pie") %>%
  layout(title = 'Race of the shooter')
```

**The Ratio of the race of the shooter**
```{r message=FALSE, warning=FALSE}
trend_race <- shoot1 %>%
  group_by(decade) %>%
  ggplot(aes(x=decade,fill=Race))+
  geom_histogram(stat="count",position="fill")+
  labs(title = "Ratio of the race")+
  xlab("Decade")+
  ylab("Ratio")
ggplotly(trend_race)
```

**Gender of the shooter**
```{r message=FALSE, warning=FALSE}
shoot1 %>%
  group_by(Gender) %>%
  count() %>%
  plot_ly(labels= ~Gender,values= ~n, type="pie") %>%
  layout(title = 'Gender of the shooter')
```

**Motive of the shooter**
```{r message=FALSE, warning=FALSE}
cause_shoot <- shoot1 %>%
  group_by(Cause) %>%
  count() %>%
  arrange(desc(n)) %>%
  ggplot(aes(x=reorder(Cause,n),y=n))+
  geom_bar(stat = 'identity',aes(fill = Cause))+
  labs(title = "Motive of the shooter")+
  xlab("Cause")+
  ylab("")+
  theme(legend.position='none')+
  coord_flip()
ggplotly(cause_shoot)
```

**Mental Status of the shooter**
```{r message=FALSE, warning=FALSE}
shoot1 %>%
  group_by(Mental.Health.Issues) %>%
  count() %>%
  plot_ly(labels= ~Mental.Health.Issues,values= ~n, type="pie") %>%
  layout(title = 'Mental Status of the shooter')
```

**Age of the shooter**
```{r message=FALSE, warning=FALSE}
age_shoot <- temp %>%
  group_by(Age) %>%
  count() %>%
  ggplot(aes(x=Age,y=n))+
  geom_bar(stat = 'identity',aes(fill = Age))+
  labs(title = "Age of the shooter")+
  xlab("Age")+
  ylab("Number of shooters")+
  theme(legend.position='none')
ggplotly(age_shoot)
```

**Age Distribution**
```{r message=FALSE, warning=FALSE}
age_dis <- shoot1 %>%
  ggplot(aes(x=decade,y=as.numeric(Age),fill=decade))+
  geom_boxplot()+
  labs(title = "Age Distribution")+
  xlab("")+
  ylab("Age")+
  theme(legend.position='none')
ggplotly(age_dis)
```


### Reference:
Samriddhi S. 2016. NYC Data Science Academy: Analyzing Mass Shooting in US. https://nycdatascience.com/blog/student-works/r-visualization/mass-shooting-us/

Bonnie Berkowitz, Denise Lu and Chris Alcantara. 2018. The Washington Post: The terrible numbers that grow with each mass shooting. https://www.washingtonpost.com/graphics/2018/national/mass-shootings-in-america/



